C{o{n{v{e{r{g{e{n{c{e}}}}}}}}}}

\href{5.html}{◀ Previous}

\hypertarget{rooted-6}{%
\section{Rooted 6}\label{rooted-6}}

\hypertarget{axiomatic-binding}{%
\subsubsection{Axiomatic binding}\label{axiomatic-binding}}

I have layed out three possible ways to create real numbers from
rationals (decimals, Dedekind, Cauchy). There are other ways, which I am
only vaguely aware of from statements in books and papers. Although they
seem different, they are found to be essentially the same through the
``axiomatic method''. The way it works is to find a set of features
(``axioms'') that the constructions share, and then prove that any
systems that share these features are mathematically the same, or
``isomorphic'' as it's called in the trade. The choice of axioms is
based on a number of criteria that include taste, minimality,
obviousness, simplicity, and so on. The criteria clash to some extent,
and the choice is often dictated by history and hence the mood of the
time of their creation, along with evolution and simplifications
according to prevailing fashion.

Most text books settle on preserving the mathematical features of the
rational numbers in terms of addition, multiplication, order, learned in
secondary school. These features can be axiomatized but they are not
``categorical'', which means that they do not lead to the systems that
satisfy them being isomorphic. In fact, there is a whole range of
``fields'', as these systems are called (in physics this word refers to
something very different), between the rationals and the reals. If the
ordering requirements are dropped, there are further fields, such as the
complex numbers and ones with just a finite number of elements (``Galois
fields''). Relaxing the multiplication axioms, so that for some elements
\(ab \neq ba\), expands the possibilities to include ``quaternions''. .
. .

The important addition for reals, as already mentioned, is the least
upper bound property. The constuctions show the existence of systems
that satisfy the proposed axioms (given only in sketch form here).
Uniqueness is shown by constucting isomorphisms between any two systems.

Firstly, to give just a sketch of what is done, there are disingished
elements, ``0'' and ``1'', in any field. It either is proved, or stated,
that these elements are unique in each particular system, so they must
be uniquely related under the isomorphism. The ``natural numbers'',
\(\mathbb N=\{0,1,2,3,\dots\}\), can be constructed uniquely in each
real system from addition of the 1s. The sequence never terminates due
to the order properties, \(n\lt n+1\), since \(0\lt 1\) and
\(x\lt y\implies n+x\lt n+y\). In similar ways, the negative integers
\(\mathbb Z^-\) and rationals \(\mathbb Q\) can be uniquely related.

The capstone is to relate the irrationals using the least upper bound
property. In any particular system, the set of rationals less than a
given element has a corresponding set of rationals in the other:
\(D(r)=\{q\in \mathbb Q: q\lt r\}\rightarrow\{q'\}\subset\mathbb Q'\).
The least upper bound of the source set is \(r\), and there is a unique
least upper bound of the target set, which we can label \(r'\), such
that \(r\rightarrow r'\), completeing the isomorphism.

A final point that may or not worry the reader is how the rationals seem
to be categorical, but their characterization as an ordered field does
not lead to this. In fact, there is an added condition that makes the
rationals categorical --- they are the \emph{minimal} ordered field or
``prime field'' contained in \(\mathbb R\).
